Implicit yaddayadda

### 2011-10-27 papers/PLDI2012/implicitization.tex
references dans la table des matieres
put title in pdf info
Revert "On ne distingue plus X_i^s et X_i^d"
This reverts commit 9dba15636aca443d1c5c5ffd5035f9f8d714ad91.
This reverts commit 3899366d305fbf8d450b66a06ff12bb04501f2f4.

\section{Introduction}
Static analysis by abstract interpretation is a fully automatic program analysis method. When applied to imperative programs, it computes an inductive invariant mapping each program location (or a subset thereof) to a set of states represented symbolically~\cite{CousotCousot_JLC92}.
For instance, if we are only interested in scalar numerical program variables, such a set may be a convex polyhedron (the set of solutions of a system of linear inequalities)~\cite{CousotHalbwachs78,PPL,BagnaraHZ08SCP}.

@@

It is a known limitation of program analysis by abstract interpretation that this convex hull, or more generally, least upper bound operation, may introduce states that cannot occur in the real program: for instance, the convex hull of the intervals $[-2,-1]$ and $[1,2]$ is $[-2,2]$, strictly larger than the union of the two.
Such introduction may prevent proving desired program properties, for instance $\neq 0$. The alternative is to keep the union symbolic (e.g. compute using $[-2,-1] \cup [1,2]$) and thus compute in the \emph{disjunctive completion}
of the lattice, but the number of terms in the union may grow exponentially with the number of successive tests in the program to analyze, not to mention difficulties for designing suitable widening operators for enforcing the convergence of fixpoint iterations~\cite{PPL,BagnaraHZ08SCP,DBLP:journals/sttt/BagnaraHZ07}.
The exponential growth of the number of terms in the union may be controlled by heuristics that judiciously apply least upper bound operations, as in the \emph{trace partitioning domain} \cite{Rival_Mauborgne_TOPLAS07} implemented in the Astr\'ee analyzer~\cite{ASTREE_PLDI03,ASTREE_ESOP05}.

Assuming we are  interested in a loop-free program fragment, the above approach of keeping symbolic unions gives the same results as performing the analysis separately over every path in the fragment.
A recent method for finding inductive loop invariants \cite{DBLP:conf/pldi/GulwaniZ10} is based on this idea: each path inside the loop body is considered separately.
Two recent proposals use SMT-solving \cite{Kroening_Strichman_08} as a decision procedure for the satisfiability of first-order arithmetic formulas in order to enumerate only paths that are needed for the progress of the analysis \cite{Gawlitza_Monniaux_ESOP11,Monniaux_Gonnord_SAS11}. They can equivalently be seen as analyses over a multigraph of transitions between some distinguished control nodes. This multigraph has an exponential number of edges, but is never explicitly represented in memory; instead, this graph is \emph{implicitly} or \emph{succinctly} represented: its edges are enumerated as needed as solutions to SMT problems.

@@

This article makes the following contributions:
\begin{enumerate}
\item We recast the guided static analysis technique from \citet{DBLP:conf/sas/GopanR07} on the expanded multigraph from \citet{Monniaux_Gonnord_SAS11}, considering entire paths instead of individual transitions, using SMT queries and binary decision diagrams.\label{contr:guided_multigraph} (See \S\ref{sec:guided_multigraph})
\item We improve the technique for obtaining disjunctive invariants from \citet{DBLP:conf/pldi/GulwaniZ10} by replacing the explicit exhaustive enumeration of paths by a sequence of SMT queries.\label{contr:disjunctive} (See \S\ref{sec:disjunctive})
\item We implemented these techniques, in addition to ``classical'' iterations and the original guided static analysis, inside a prototype static analyzer.
This tool uses the LLVM bitcode format \cite{Lattner:2004:LCF:977395.977673,LLVM_langref} as input, which can be produced by compilation from C, C++ and Fortran, enabling it to be run on many real-life programs.
With respect to abstract domains on numerical variables, it uses the APRON library \cite{DBLP:conf/cav/JeannetM09}, which supports a variety of abstract domains from which we can choose with minimal changes to our analyzer.
This article summarizes our implementation decisions for analysis of SSA-form programs.
\item We conducted extensive experiments with this tool, on real-life programs, thus obtaining experimental comparisons between the various combinations of techniques. (\S\ref{sec:experiments})
\end{enumerate}

@@

Once an inductive invariant $\gamma((\abstr{x}_p)_{p \in P})$ has been obtained, one can attempt \emph{decreasing} or \emph{narrowing} iterations to reduce it. In their simplest form, this just means running the following operation until a fixpoint or a maximal number of iterations are reached: for any $p'$, replace $\abstr{x}_{p'}$ by $\abstr{x}_p \cap \left(\bigsqcup_{p \in P} \abstr{\tau}_{p,p'} (\abstr{x}_p)\right)$. The result also defines an inductive invariant. These decreasing iterations are indispensable to recover properties from guards (tests) in the program in most iteration settings; unfortunately, certain loops, particularly those involving identity (no-operation) transitions, may foil them: the iterations immediately reach a fixpoint and do not decrease further (see example in \S\ref{subsec:rate_lim}). Sections \ref{sec:guided} and \ref{sec:path_focusing} describe techniques that work around this problem.

@@

The expected inductive invariant is $\verb|x_old| \in [-100000,100000]$, but classical abstract interpretation using intervals (or octagons or polyhedra) finds $\verb|x_old| \in (-\infty,+\infty)$~\citep{ASTREE_ESOP05}.
Let us briefly see why.

Widening iterations converge to $\lstinline|x_old| \in (-\infty,+\infty)$; let us now see why decreasing iterations fail to recover the desired invariant.
The \lstinline|x >= x_old+10| test at line~6, if taken, yields $\lstinline|x_old| \in (-\infty,99990)$; followed by \lstinline|x = x_old+10|, we obtain $\lstinline|x| \in (-\infty,\allowbreak 100000]$, and the same after union with the no-operation ``else'' branch. Line~7 yields $\lstinline|x| \in (-\infty,+\infty)$.

@@

\subsection{Guided static analysis}
\label{sec:guided}
\emph{Guided static analysis} was proposed by \citet{DBLP:conf/sas/GopanR07} as an improvement over classical upward Kleene iterations with widening.
Consider the following program:
\lstinputlisting[label=lst:gopan_reps]{gopan_reps.c}

\begin{figure}
\begin{center}
\includegraphics{figures/gopan_reps}
\end{center}
\caption{The invariant for Prog.~\ref{lst:gopan_reps}: the piecewise linear, solid line is the strongest invariant, the grayed polyhedron is its convex hull.}
\label{fig:gopan_reps_invariant} 
\end{figure}

Classical iterations on the domain of convex polyhedra \cite{Halbwachs_PhD,CousotHalbwachs78} or octagons \cite{Mine_PhD,DBLP:journals/lisp/Mine06} start with $x = 0 \land x = 0$, then continue with $x = y \land 0 \leq x \leq 1$.
The widening operator extrapolates from these two iterations and yields $x = y \land x \geq 0$.
From there, the ``else'' branch at line 5 may be taken; with further widening, $0 \leq y \leq x$ is obtained as a loop invariant, and thus the postcondition computed at line 9 is $x \geq 0 \land y = 0$.
Yet the strongest invariant is $(0 \leq x \leq 51 \land y = x) \lor (51 \leq x \leq 102 \land x+y=102)$, and its convex hull, a convex polyhedron (Fig.~\ref{fig:gopan_reps_invariant}), is
\begin{equation}
y \leq x \land y \leq 102-x \land y \geq 0.\label{eqn:triangle}
\end{equation}
Intuitively, this disappointing result is obtained because widening extrapolates from the first iterations of the loop, but the loop has two different phases ($x \leq 50$ and $x > 50$) with different behaviors, thus the extrapolation from the first phase is not valid for the second.

@@

Guided static analysis \cite{DBLP:conf/sas/GopanR07} applies to the transition
graph of the program. We now present a new technique applying this analysis on the implicit
multigraph from \cite{Monniaux_Gonnord_SAS11}, thus avoiding control flow merges with
unfeasible paths.
In this section, we use the same notations as \ref{sec:path_focusing}.

@@

In order to find which path to focus on, we construct an SMT formula $f(p_i)$, whose
model when satisfiable is a path that starts in $p_i$, goes to a successor $p_j
\in P_R$ of $p_i$, such that the image of $X_{i}$ by the path transformation
is not included in the current $X_{j}$.
Intuitively, such a path makes the abstract value $X_{j}$ grow, and thus is
an interesting path to focus on. We loop until the formula becomes unsatisfiable,
meaning that the analysis of $p_i$ is finished.

If we note $Succ(i)$ the set of indices $j$ such that $p_j \in P_R$ is a
successor of $p_i$ in the expanded multigraph, and $X_i$ the abstract value
associated to $p_i$ :
$$f(p_i) = \rho \wedge b_i^s \wedge 
\displaystyle\bigwedge_{j \in P_R \atop j \neq i} \neg
b_j^s \wedge x \in X_i \wedge \displaystyle\bigvee_{j \in Succ(i)} (b_j^d \wedge
x \notin X_j)$$

@@

@@

This motivates a key implement decision of our tool: only those variables $v_1,\dots,v_n$ that are not defined by arithmetic operations are retained as coordinates in the abstract domain (e.g. as dimensions in polyhedra). A basic block of code therefore amounts to a \emph{parallel assignment} operation
$(v_1,\dots,v_n) \allowbreak\mapsto\allowbreak
(f_1(v_1,\dots,v_n), \allowbreak, \dots, \allowbreak
 f_n(v_1,\dots,v_n))$;
such operations are directly supported by APRON.
A first benefit is to limit the number of dimensions in the abstract values, since polyhedra libraries typically perform worse with higher dimensions.
A second benefit is that the abstract operation for a single path in path-focusing methods also is a (large) parallel assignment.

We further reduce the number of variables by projecting out those that are not live at the program point. Again, SSA use-def and def-use chains make it easy to compute liveness.

@@

Our implementation of path-focusing currently does not use true acceleration techniques, as proposed by \citet{Monniaux_Gonnord_SAS11}. Instead, it simply runs widening and narrowing iterations on a single path.

>>>>

\section{Introduction}
% TODO expliquer 'implicite'
Static analysis by abstract interpretation is a fully automatic program analysis method. When applied to imperative programs, it computes an inductive invariant mapping each program location (or a subset thereof) to a set of states represented symbolically~\cite{CousotCousot_JLC92}.
For instance, if we are only interested in scalar numerical program variables, such a set may be a convex polyhedron (the set of solutions of a system of linear inequalities)~\cite{CousotHalbwachs78,PPL,BagnaraHZ08SCP}.

@@

It is a known limitation of program analysis by abstract interpretation that this convex hull, or more generally, least upper bound operation, may introduce states that cannot occur in the real program: for instance, the convex hull of the intervals $[-2,-1]$ and $[1,2]$ is $[-2,2]$, strictly larger than the union of the two.
Such introduction may prevent proving desired program properties, for instance $\neq 0$. The alternative is to keep the union symbolic (e.g. compute using $[-2,-1] \cup [1,2]$) and thus compute in the \emph{disjunctive completion}
of the lattice, but the number of terms in the union may grow exponentially with the number of successive tests in the program to analyze, not to mention difficulties for designing suitable widening operators for enforcing the convergence of fixpoint iterations~\cite{PPL,BagnaraHZ08SCP,DBLP:journals/sttt/BagnaraHZ07}.
The exponential growth of the number of terms in the union may be controlled by heuristics that judiciously apply least upper bound operations, as in the \emph{trace partitioning domain} \cite{Rival_Mauborgne_TOPLAS07} implemented in the Astr\'ee analyzer~\cite{ASTREE_TASE07,ASTREE_ESOP05,ASTREE_PLDI03}. % TODO ne garder que les papiers les plus significatifs par rapport au sujet évoqué

Assuming we are  interested in a loop-free program fragment, the above approach of keeping symbolic unions gives the same results as performing the analysis separately over every path in the fragment.
A recent method for finding inductive loop invariants \cite{DBLP:conf/pldi/GulwaniZ10} is based on this idea: each path inside the loop body is considered separately.
Two recent proposals use SMT-solving \cite{Kroening_Strichman_08} as a decision procedure for the satisfiability of first-order arithmetic formulas in order to enumerate only paths that are needed for the progress of the analysis \cite{Gawlitza_Monniaux_ESOP11,Monniaux_Gonnord_SAS11}.

@@

This article makes the following contributions:
\begin{enumerate}
\item We recast the guided static analysis technique from \citet{DBLP:conf/sas/GopanR07} on the expanded multigraph from \citet{Monniaux_Gonnord_SAS11}, considering entire paths instead of individual transitions, using SMT queries and binary decision diagrams.\label{contr:guided_multigraph} (See \S\ref{sec:guided_multigraph})
\item We improve the technique for obtaining disjunctive invariants from \citet{DBLP:conf/pldi/GulwaniZ10} by replacing the explicit exhaustive enumeration of paths by a sequence of SMT queries.\label{contr:disjunctive} (See \S\ref{sec:disjunctive})
% TODO: 'We implemented' pas une contribution; rephrase.
\item We implemented these techniques, in addition to ``classical'' iterations and the original guided static analysis, inside a prototype static analyzer.
This tool uses the LLVM bitcode format \cite{Lattner:2004:LCF:977395.977673,LLVM_langref} as input, which can be produced by compilation from C, C++ and Fortran, enabling it to be run on many real-life programs.
With respect to abstract domains on numerical variables, it uses the APRON library \cite{DBLP:conf/cav/JeannetM09}, which supports a variety of abstract domains from which we can choose with minimal changes to our analyzer.
\item We conducted extensive experiments with this tool, on real-life programs, thus obtaining experimental comparisons between the various combinations of techniques. (\S\ref{sec:experiments})
\end{enumerate}

@@

@@

The expected inductive invariant is $\verb|x_old| \in [-100000,100000]$, but classical abstract interpretation using intervals (or octagons or polyhedra) finds $\verb|x_old| \in (-\infty,+\infty)$. % TODO chercher ref Astree
Let us briefly see why.

Widening iterations converge to $\lstinline|x_old| \in (-\infty,+\infty)$; let us now see why decreasing iterations fail to recover the desired invariant. % TODO expliquer widening suivi de narrowing
The \lstinline|x >= x_old+10| test at line~6, if taken, yields $\lstinline|x_old| \in (-\infty,99990)$; followed by \lstinline|x = x_old+10|, we obtain $\lstinline|x| \in (-\infty,\allowbreak 100000]$, and the same after union with the no-operation ``else'' branch. Line~7 yields $\lstinline|x| \in (-\infty,+\infty)$.

@@

\subsection{Guided static analysis}
\label{sec:guided}
\emph{Guided static analysis} was proposed by \citet{DBLP:conf/sas/GopanR07} as an improvement over classical upward Kleene iterations with widening.
Consider the following program:
\lstinputlisting{gopan_reps.c}

Classical iterations on the domain of convex polyhedra \cite{Halbwachs_PhD,CousotHalbwachs78} or octagons \cite{Mine_PhD,DBLP:journals/lisp/Mine06} start with $x = 0 \land x = 0$, then continue with $x = y \land 0 \leq x \leq 1$.
The widening operator extrapolates from these two iterations and yields $x = y \land x \geq 0$.
From there, the ``else'' branch at line 5 may be taken; with further widening, $0 \leq y \leq x$ is obtained as a loop invariant, and thus the postcondition computed at line 9 is $x \geq 0 \land y = 0$.
Yet the strongest invariant is $(0 \leq x \leq 51 \land y = x) \lor (51 \leq x \leq 102 \land x+y=102)$, and its convex hull, a convex polyhedron, is
\begin{equation}
y \leq x \land y \leq 102-x \land y \geq 0.\label{eqn:triangle}
\end{equation} % TODO fig
Intuitively, this disappointing result is obtained because widening extrapolates from the first iterations of the loop, but the loop has two different phases ($x \leq 50$ and $x > 50$) with different behaviors, thus the extrapolation from the first phase is not valid for the second.

@@

Guided static analysis \cite{DBLP:conf/sas/GopanR07} applies to the transition
graph of the program. We now present a new technique applying this analysis on the implicit
multigraph from \cite{Monniaux_Gonnord_SAS11}, thus avoiding control flow merges with
unfeasible paths.
In this section, we use the same notations as \ref{sec:path_focusing}, except
that we call the abstract values $X_i^s$ instead of $X_i$ to leave room for
another abstract value $X_i^d$ later in the algorithm.

@@

In order to find which path to focus on, we construct an SMT formula $f(p_i)$, whose
model when satisfiable is a path that starts in $p_i$, goes to a successor $p_j
\in P_R$ of $p_i$, such that the image of $X_{i}^s$ by the path transformation
is not included in the current $X_{j}^s$.
Intuitively, such a path makes the abstract value $X_{j}^s$ grow, and thus is
an interesting path to focus on. We loop until the formula becomes unsatisfiable,
meaning that the analysis of $p_i$ is finished.

If we note $Succ(i)$ the set of indices $j$ such that $p_j \in P_R$ is a
successor of $p_i$ in the expanded multigraph, and $X_i^s$ the abstract value
associated to $p_i$ :
$$f(p_i) = \rho \wedge b_i^s \wedge 
\displaystyle\bigwedge_{j \in P_R \atop j \neq i} \neg
b_j^s \wedge x \in X_i^s \wedge \displaystyle\bigvee_{j \in Succ(i)} (b_j^d \wedge
x \notin X_j^s)$$

@@

We use another
abstract value associated to a control point $p_j$, noted $X_j^d$, which is
distinct from $X_j^s$, and initialized to $X_j^s$.

@@

This motivates a key implement decision of our tool: only those variables $v_1,\dots,v_n$ that are not defined by arithmetic operations are retained as coordinates in the abstract domain (e.g. as dimensions in polyhedra). A basic block of code therefore amounts to a \emph{parallel assignment} operation
$(v_1,\dots,v_n) \allowbreak\mapsto\allowbreak
(f_1(v_1,\dots,v_n), \allowbreak, \dots, \allowbreak
 f_n(v_1,\dots,v_n))$;
such operations are directly supported by APRON.
A first benefit is to limit the number of dimensions in the abstract values, since polyhedra libraries typically perform worse with higher dimensions. TODO
Note that this means that the abstract operation for a single path in path-focusing methods also is a (large) parallel assignment.

@@
